Exploring probability distributions for bivariate temporal granularities

### 2020-01-30 main.Rmd
commit after long delays of Rstudio

The variable `reading_datetime` is the index of the tsibble and represents the time variable. The variable `customer_id` denoting different households represent the key of the tsibble and `general_supply_kwh` is the time series variable which needs to be analyzed. Data sets of the form <C1, C2, `general_supply_kwh`> will form the basis of exploration for each key in the data set, where the cyclic granularity C1 would be mapped to x axis and `general_supply_kwh` to y-axis with the cyclic granularity C2 mapped to facet. Any geometry displaying the probability distribution of `general_supply_kwh` like  boxplot, letter value, violin,
ridge or highest density region plots can then be used to explore the data set.

@@

smart_meter <- sm %>% as_tsibble()

smart_meter %>% kable()
```

@@

```{r, echo=TRUE}
library(gravitas)
smart_meter %>% search_gran()
```

If these options are considered too many, the most coarse temporal unit can be set to be a “month”.

```{r, echo=TRUE}
smart_meter10 %>% 
  search_gran(highest_unit = "month")
```

@@

smart_meter10 %>% search_gran(highest_unit = "month",
                              filter_out = c("hhour", "fortnight")
)
```

@@

From the search list, we found six cyclic granularities for which we would like to derive insights of energy behavior. Given the data structure <$C1$ $C2$, `general_supply_kwh`>, each of those six cyclic granularities can either be mapped to x-axis or to facet. Thus the problem is equivalent to taking 2 granularities at a time from 6, which essentially is equivalent to visualizing 30 plots. However, harmony/clash pairs can be identified among those 30 pairs, to determine feasibility of plotting any pairs together. We are left with 13 harmonies pair, each of which can be plotted together to look at the energy behavior from different perspectives.

@@

```{r harmony, echo=TRUE}
smart_meter10 %>% harmony(
  ugran = "month",
  filter_out = c("hhour", "fortnight")
)
smart_meter10 %>% gran_advice("wknd_wday", "hour_day")

@@

However, if the data for all keys are visualized together, it might lead to Simpson's paradox, which occurs when one `customer_id` shows a particular behavior, but this behavior is reversed when all keys are combined together. This is also intuitive because `customer_id`'s with very different occupation or demographics will tend to have very different energy behavior and combining them together will somehow weaken any typical or extreme behavior. For example, Figure \ref{fig:wknd-wday50} shows the quantile plot of energy consumption of a sample of 50 households. Quantiles for weekend and weekdays for these customers look exactly the same. It is not clear if the behaviors of most of these customers do not alter between weekdays and weekends or looking at all customers together balances out the behaviors. Consider another example in Figure \ref{fig:my-hd} showing the area quantile plots of 50 customers. The median is very close to the lower boundaries of all the other bands implying energy consumption for these households are left skewed. The pink band changes significantly in winter months (May - August) implying that much of the behavioral changes occur in the quartiles, that too in peak hours of the day. It is to be noted here that the the level of all quantiles increased too in winter months (increased energy usage in weather conditions), the interesting part is to notice that the relationship between bands other than quartile stayed same across seasons.

@@

Since we established that looking at random samples with different demographics might confound our analysis, next the smart meter data is filtered for two customers, to illustrate kinds of insights that can be drawn for the energy behavior of these two customers

@@

In \autoref{fig:bothcust} we visualize the harmony pair (wknd_wday, hour_day) through a box plot. Boxplot of energy consumption is shown across `wknd_wday` (facet) and `hour_day` (x-axis) for the same two households. For the first household, the shape of the distribution does not change significantly for weekdays and weekends. The distribution is roughly smooth with a morning and an evening peak. For the second household, the quantiles behave differently in peak hours. For the first few hours on weekdays, energy consumption starts at a much higher level than weekends. Consider looking at letter value plots instead of quantile plots to look at the same data in \autoref{fig:bothcust-lv}. There is additional information that we can derive looking at the distribution.
There is bimodality in the early hours of the day, implying both low and high energy consumption is probable in the early hours of the day. Also hours from 7 to 13 seems most volatile given the shape of the violin. If we visualize the same data with reverse mapping of the cyclic granularities, then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. For example in \autoref{fig:cust-lv}, it can be seen that median energy consumption for the early morning hours is extremely high for weekdays compared to weekends. Also, outliers are more prominent in the latter part of the day. All of these indicates that looking at different distribution plots or changing the mapping might draw lights on different aspect of the energy behavior for the same customer.

@@

This case study shows systematic exploration of energy behavior for two random households comparing some of their features with all 50 households. First, it helps us to find the list of granularities to look at, then shrinks the number of possible visualizations by identifying harmonies and lastly visualise harmony pairs to gain some insights on periodic behavior of the households.

>>>>

The variable `reading_datetime` is the index of the tsibble and represents the time variable. The variable `customer_id` denoting different households represent the key of the tsibble and `general_supply_kwh` is the time series variable which needs to be analyzed. Data sets of the form <$C1$, $C2$, `general_supply_kwh`> will form the basis of exploration for each key in the data set, where the cyclic granularity $C1$ would be mapped to $x$ axis and `general_supply_kwh` to $y$ axis with the cyclic granularity $C2$ mapped to facet. Any geometry displaying the probability distribution of `general_supply_kwh` like  boxplot, letter value, violin,
ridge or highest density region plots can then be used to explore the data set.

@@

smart_meter <- sm %>% as_tsibble() %>% dplyr::select(customer_id, 
                              reading_datetime, 
                              general_supply_kwh, 
                              everything()) %>% head()

knitr::kable(smart_meter, format = "markdown")
```

@@

```{r search, echo=TRUE}
library(gravitas)
smart_smart_search <- smart_meter %>% search_gran()
knitr::kable(smart_smart_search, format = "markdown")
```

If these options are considered too many, the most coarse temporal unit can be set to a “month”.

```{r search_limit1, echo=TRUE}
smart_meter_limit1 <- smart_meter10 %>% 
  search_gran(highest_unit = "month")
knitr::kable(smart_meter_limit1, format = "markdown")
```

@@

smart_meter_limit2 <- smart_meter10 %>% search_gran(highest_unit = "month",
                              filter_out = c("hhour", "fortnight")
)
knitr::kable(smart_meter_limit2, format = "markdown")
```

@@

From the search list, we found six cyclic granularities for which we would like to derive insights of energy behavior. Given the data structure <$C1$ $C2$, `general_supply_kwh`>, each of those six cyclic granularities can either be mapped to x-axis or to facet. Thus the problem is equivalent to taking 2 granularities at a time from 6, which essentially is equivalent to having 30 data subsets for visualisation. However, harmony/clash pairs can be identified among those 30 possibilties to determine feasibility of plotting any pairs together. We are left with 13 harmonies pair, each of which can be plotted together to look at the energy behavior from different perspectives.

@@

```{r harmony, echo=TRUE}
smart_meter10 %>% harmony(
  ugran = "month",
  filter_out = c("hhour", "fortnight")
)
# smart_meter10 %>% gran_advice("wknd_wday", "hour_day")

@@

If the data for all keys are visualized together, it might lead to Simpson's paradox, which occurs when one `customer_id` shows a particular behavior, but this behavior is reversed when all keys are combined together. This is also intuitive because `customer_id`'s with very different occupation or demographics will tend to have very different energy behavior and combining them together will somehow weaken any typical or extreme behavior. For example, Figure \ref{fig:wknd-wday50} shows the quantile plot of energy consumption of a sample of 50 households. Quantiles for weekends and weekdays for these customers look exactly the same. It is not clear if the behaviors of most of these customers do not alter between weekdays and weekends or looking at all customers together balances out the behaviors. Consider another example in Figure \ref{fig:my-hd} showing the area quantile plots of 50 customers. The median is very close to the lower boundaries of all the other bands implying energy consumption for these households are left skewed. The pink band changes significantly in winter months (May - August) implying that much of the behavioral changes occur in the quartiles, that too in peak hours of the day. It is to be noted here that the the level of all quantiles increased too in winter months (increased energy usage in weather conditions), the interesting part is to notice that the relationship between bands other than quartile stayed same across seasons.

@@

Looking at random samples with different demographics might confound our analysis and a strategy for analyzing multiple keys together could be to first screen some basis time series or demographic features and then look at their energy behavior. This is beyond the scope of the current work. For the purpose of this paper, the smart meter data is filtered for two customers to illustrate kinds of insights that can be drawn after proceeding with the systematic exploration.

@@

In \autoref{fig:bothcust} we visualize the harmony pair (wknd_wday, hour_day) through a box plot. Boxplot of energy consumption is shown across `wknd_wday` (facet) and `hour_day` (x-axis) for the same two households. For the first household, the shape of the distribution does not change significantly for weekdays and weekends. The distribution is roughly smooth with a morning and an evening peak. For the second household, the quantiles behave differently in peak hours. For the first few hours on weekdays, energy consumption starts at a much higher level than weekends. Consider looking at letter value plots instead of quantile plots to look at the same data in \autoref{fig:bothcust-lv}. There is additional information that we can derive looking at the distribution.
There is bimodality in the early hours of the day, implying both low and high energy consumption is probable in the early hours of the day. Also hours from 7 to 13 look most volatile. If we visualize the same data with reverse mapping of the cyclic granularities, then the natural tendency would be to compare weekend and weekday behavior within each hour and not across hours. For example in \autoref{fig:cust-lv}, it can be seen that median energy consumption for the early morning hours is extremely high for weekdays compared to weekends. Also, outliers are more prominent in the latter part of the day. All of these indicates that looking at different distribution plots or changing the mapping might draw lights on different aspect of the energy behavior for the same customer.

@@

This case study shows systematic exploration of energy behavior for two random households to gain some insights on periodic behavior of the households. First, it helps us to find the list of cyclic granularities to look at, then shrinks the number of possible visualizations by identifying harmonies, visualise a harmony pair and shows the effect of different distribution plots or reverse mapping.
